---
title: "Logging with logger: namespaces and thresholds"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Logging with logger: namespaces and thresholds}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

**tooltipexplorer** uses the [`logger`](https://daroczig.github.io/logger/)
package for structured, namespace-aware logging throughout the application.
Every log call carries a `namespace` argument that identifies exactly which
module or layer emitted the message, making it straightforward to filter noise
in development and route messages selectively in production.

---

## Log-level hierarchy

`logger` defines seven levels from most to least verbose:

| Level | Function | When to use |
|---|---|---|
| `TRACE` | `log_trace()` | Fine-grained internal steps (loops, branches) |
| `DEBUG` | `log_debug()` | Per-reactive / per-call diagnostics |
| `INFO` | `log_info()` | Key lifecycle events (session start, fetch, render complete) |
| `SUCCESS` | `log_success()` | Explicit success confirmations |
| `WARN` | `log_warn()` | Recoverable issues (empty ticker list, unexpected input) |
| `ERROR` | `log_error()` | Caught errors before re-throwing |
| `FATAL` | `log_fatal()` | Unrecoverable failures |

The package default threshold is `INFO`: `DEBUG` and `TRACE` lines are silent
unless you explicitly lower the threshold.

---

## Namespaces in this package

Every `logger::log_*()` call in **tooltipexplorer** passes an explicit
`namespace` string.  The full set, registered in `app_set_log_threshold()`, is:

| Namespace | File | Covers |
|---|---|---|
| `"global"` | — | logger's built-in global namespace (fallback) |
| `"tooltipexplorer/app"` | `app_ui.R`, `app_server.R` | UI construction, session lifecycle, module wiring |
| `"tooltipexplorer/inputs"` | `mod_inputs.R` | Fetch-button events, reactive inputs list |
| `"tooltipexplorer/outputs"` | `mod_outputs.R` | Price fetch, returns, performance, all render calls |
| `"tooltipexplorer/download"` | `mod_download.R` | Filename generation, report render |
| `"tooltipexplorer/tooltip"` | `mod_tooltip.R` | Tooltip helper dispatch |
| `"tooltipexplorer/hoverinfo"` | `mod_hoverinfo.R` | Hover-span construction |

The `"tooltipexplorer/<module>"` convention means you can silence one noisy
module while keeping the others verbose — see the examples below.

---

## `app_set_log_threshold()`

This is the single entry point for changing thresholds.  It is called once per
session in `app_server()` and applies the same level to every namespace in the
table above.

```{r app_set_log_threshold-source}
app_set_log_threshold <- function(level = logger::INFO) {
  namespaces <- c(
    "global",
    "tooltipexplorer/app",
    "tooltipexplorer/inputs",
    "tooltipexplorer/outputs",
    "tooltipexplorer/download",
    "tooltipexplorer/tooltip",
    "tooltipexplorer/hoverinfo"
  )
  for (ns in namespaces) {
    logger::log_threshold(level, namespace = ns)
  }
  invisible(level)
}
```

### Common threshold recipes

```{r threshold-recipes}
# Default (production) — INFO and above only
tooltipexplorer::app_set_log_threshold(logger::INFO)

# Development — everything including DEBUG
tooltipexplorer::app_set_log_threshold(logger::DEBUG)

# Silent except for warnings and errors
tooltipexplorer::app_set_log_threshold(logger::WARN)
```

You can call `app_set_log_threshold()` interactively at any time while the app
is running — the threshold change takes effect immediately for all subsequent
log calls.

---

## Per-namespace overrides

After calling `app_set_log_threshold()` you can override individual namespaces
with `logger::log_threshold()` directly:

```{r per-ns-override}
# Make the outputs module verbose while keeping everything else at INFO
tooltipexplorer::app_set_log_threshold(logger::INFO)
logger::log_threshold(logger::DEBUG, namespace = "tooltipexplorer/outputs")

# Silence the download module entirely
logger::log_threshold(logger::FATAL, namespace = "tooltipexplorer/download")
```

### Reading back the current threshold

```{r read-threshold}
# Returns the current level for a given namespace
logger::log_threshold(namespace = "tooltipexplorer/outputs")
```

---

## `with_logging()`

`with_logging()` is a `tryCatch` / `withCallingHandlers` wrapper used around
expressions that might produce warnings or errors.  It logs both, adds a
`context` label to every message, and then re-issues the condition so Shiny
and the caller continue to see it normally.

```{r with_logging-source}
with_logging <- function(expr, context = "", ns = "tooltipexplorer/app") {
  tryCatch(
    withCallingHandlers(
      expr,
      warning = function(w) {
        logger::log_warn(
          "[{context}] Warning: {conditionMessage(w)}",
          namespace = ns
        )
        invokeRestart("muffleWarning")
      }
    ),
    error = function(e) {
      logger::log_error(
        "[{context}] Error: {conditionMessage(e)}",
        namespace = ns
      )
      stop(e)
    }
  )
}
```

`with_logging()` is used in three places in `app_server()`:

```{r with_logging-in-app-server}
# Wiring the inputs module
inputs_r <- with_logging(
  mod_inputs_server("inputs"),
  context = "app_server / mod_inputs_server",
  ns      = "tooltipexplorer/app"
)

# Wiring the outputs module
perf_r <- with_logging(
  mod_outputs_server("outputs", inputs_r = inputs_r),
  context = "app_server / mod_outputs_server",
  ns      = "tooltipexplorer/app"
)

# Wiring the download module
with_logging(
  mod_download_server("download", inputs_r = inputs_r, perf_r = perf_r),
  context = "app_server / mod_download_server",
  ns      = "tooltipexplorer/app"
)
```

And inside module server functions themselves, wrapping individual output
renderers:

```{r with_logging-in-renderUI}
output$bslib_boxes <- shiny::renderUI({
  shiny::req(perf_r())
  with_logging(
    context = "mod_outputs / bslib_boxes",
    ns      = "tooltipexplorer/outputs",
    {
      df <- perf_r()
      logger::log_debug(
        "Rendering bslib popover boxes | n: {nrow(df)}",
        namespace = "tooltipexplorer/outputs"
      )
      # ... render code ...
    }
  )
})
```

---

## Log-level patterns by module

### `app_ui()` and `app_server()` — namespace `"tooltipexplorer/app"`

| Event | Level |
|---|---|
| `"Building app UI"` | `INFO` |
| `"Session started | session_id: ..."` | `INFO` |
| `"mod_inputs_server() ready"` | `INFO` |
| `"mod_outputs_server() ready"` | `INFO` |
| `"mod_download_server() ready"` | `INFO` |
| `"Session ended | session_id: ..."` | `INFO` |
| Any warning/error from `with_logging()` | `WARN` / `ERROR` |

### `mod_inputs_server()` — namespace `"tooltipexplorer/inputs"`

| Event | Level |
|---|---|
| `"mod_inputs_server() initialised | id: ..."` | `DEBUG` |
| `"Fetch button pressed | tickers: [...] | from: ... | to: ... | vol_window: ..."` | `INFO` |
| `"Fetch pressed with no tickers selected"` | `WARN` |
| `"Inputs reactive evaluated | tickers: [...]"` | `DEBUG` |

### `mod_outputs_server()` — namespace `"tooltipexplorer/outputs"`

| Event | Level |
|---|---|
| `"mod_outputs_server() initialised | id: ..."` | `DEBUG` |
| `"Fetching prices | tickers: [...] | from: ... | to: ..."` | `INFO` |
| `"Price fetch failed | tickers: [...] | error: ..."` | `ERROR` |
| `"Prices fetched | rows: ... | tickers: [...]"` | `INFO` |
| `"Computing daily returns"` | `DEBUG` |
| `"get_stock_returns() failed | error: ..."` | `ERROR` |
| `"Returns computed | rows: ..."` | `DEBUG` |
| `"Computing performance summary"` | `DEBUG` |
| `"summarise_performance() failed | error: ..."` | `ERROR` |
| `"Performance summary ready | symbols: [...]"` | `INFO` |
| `"Rendering value boxes | n: ..."` | `DEBUG` |
| `"Rendering bslib popover boxes | n: ..."` | `DEBUG` |
| `"Rendering shinyhelper cards | n: ..."` | `DEBUG` |
| `"Rendering prompter cards | n: ..."` | `DEBUG` |
| `"Rendering shinyalert cards | n: ..."` | `DEBUG` |
| `"Rendering reactable table | n: ..."` | `DEBUG` |

### `mod_download_server()` — namespace `"tooltipexplorer/download"`

| Event | Level |
|---|---|
| `"mod_download_server() initialised | id: ..."` | `DEBUG` |
| `"Download filename generated | file: ..."` | `INFO` |
| `"Report render started | format: ... | tickers: [...]"` | `INFO` |
| `"report_template.Rmd not found in package inst/"` | `ERROR` |
| `"Rendering to temp dir | path: ..."` | `DEBUG` |
| `"rmarkdown::render() failed | format: ... | error: ..."` | `ERROR` |
| `"Report render complete | format: ... | file: ..."` | `INFO` |

### `mod_hoverinfo()` — namespace `"tooltipexplorer/hoverinfo"`

| Event | Level |
|---|---|
| `"mod_hoverinfo() | type: ... | contents length: ..."` | `DEBUG` |
| `"mod_hoverinfo() building reactable span | named: ..."` | `DEBUG` |
| `"mod_hoverinfo() failed | type: ... | error: ..."` | `ERROR` |

---

## Log message format

All messages use `logger`'s
[`glue`](https://glue.tidyverse.org/)-interpolation syntax — curly-brace
expressions are evaluated in the calling environment at the time the log call
is made:

```{r message-format}
# Structured fields separated by " | "
logger::log_info(
  "Fetching prices | tickers: [{paste(tickers, collapse = ', ')}] | from: {from} | to: {to}",
  namespace = "tooltipexplorer/outputs"
)
# → INFO [2026-04-03 08:00:00] Fetching prices | tickers: [AAPL, MSFT] | from: 2024-01-01 | to: 2024-12-31
```

The `key: value` pattern (pipe-separated) makes it easy to `grep` for a
specific field across a log file:

```bash
grep "tickers:" app.log
grep "ERROR" app.log
grep "tooltipexplorer/outputs" app.log
```

---

## Writing to a file

By default `logger` writes to the console.  To add a file appender alongside
the console appender:

```{r file-appender}
# Append all INFO+ messages to a rotating log file
logger::log_appender(
  logger::appender_tee(file = "tooltipexplorer.log"),
  namespace = "tooltipexplorer/app"
)
```

Or route a specific namespace to its own file:

```{r file-appender-per-ns}
logger::log_appender(
  logger::appender_file("outputs.log"),
  namespace = "tooltipexplorer/outputs"
)
```

---

## Quick-reference: development setup

```{r dev-setup}
# 1. Load the package
devtools::load_all()

# 2. Make all namespaces verbose
tooltipexplorer::app_set_log_threshold(logger::DEBUG)

# 3. (Optional) focus on one noisy namespace only
logger::log_threshold(logger::DEBUG, namespace = "tooltipexplorer/outputs")

# 4. Launch — log messages stream to the console as you interact
tooltipexplorer::launch()
```

---

## References

- [logger package documentation](https://daroczig.github.io/logger/)
- `?tooltipexplorer::app_set_log_threshold`
- `?tooltipexplorer::with_logging`
