---
title: "Tooltip and Hover-Info Modules"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tooltip and Hover-Info Modules}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

Two helper functions — `mod_tooltip()` and `mod_hoverinfo()` — give every
module in **tooltipexplorer** a consistent, package-agnostic way to attach
tooltip and hover-info content to UI elements and outputs.

- **`mod_tooltip()`** is a *UI helper*. It returns a `shiny.tag` and can be
  placed anywhere in a UI tree or inside another module's UI function — no
  server-side counterpart is needed.
- **`mod_hoverinfo()`** is a *server/logic helper*. It formats hover content
  for back-ends that build their tooltips programmatically (reactable, gt) rather than through HTML attributes.

The two functions cover complementary ground:

| Question | Function |
|---|---|
| "I want a hover/click tooltip on a *label* or *icon* in the UI" | `mod_tooltip()` |
| "I want hover content inside a *chart or table cell*" | `mod_hoverinfo()` |

---

## `mod_tooltip()`

### Signature

```{r}
mod_tooltip(
  trigger  = bsicons::bs_icon("info-circle"),
  type     = c("bslib", "shinyhelper", "prompter", "shinyalert"),
  contents = "",
  size     = NULL,
  style    = NULL,
  ...
)
```

### Parameters

| Parameter | Type | Description |
|---|---|---|
| `trigger` | `shiny.tag` or text | The element the tooltip attaches to. Defaults to a small info-circle icon. |
| `type` | character | Back-end: `"bslib"` (default), `"shinyhelper"`, `"prompter"`, or `"shinyalert"`. |
| `contents` | character | Tooltip/popover/modal body text. HTML is accepted for all back-ends. |
| `size` | character | CSS font-size for the trigger wrapper, e.g. `"0.85rem"`. `NULL` leaves it unchanged. |
| `style` | character | Additional inline CSS for the trigger wrapper, e.g. `"color:#6c757d"`. `NULL` applies none. |
| `...` | — | Extra arguments forwarded to the underlying back-end function. See per-back-end notes below. |

### How it works

`size` and `style` are combined into a single inline CSS string and applied to
a `<span>` that wraps `trigger`. The wrapped trigger is then passed to the
chosen back-end via `switch()`.

```
mod_tooltip(trigger, type, contents, size, style, ...)
       │
       ├─ build inline CSS from size + style
       ├─ wrap trigger in <span style="..."> if CSS is non-NULL
       │
       └─ switch(type)
            ├─ "bslib"       → bslib::popover(trigger, HTML(contents), ...)
            ├─ "shinyhelper" → shinyhelper::helper(trigger, content = contents, ...)
            ├─ "prompter"    → prompter::add_prompt(trigger, message = contents, ...)
            └─ "shinyalert"  → tagList(useShinyalert(), <span onclick="shinyalert(...)">)
```

### Usage in a module UI

The most common use is attaching a small info icon to an input label so the
label carries its own contextual help without any server code:

```{r}
# Inside mod_inputs_ui() — bslib popover on a slider label
shiny::sliderInput(
  inputId = ns("vol_window"),
  label   = shiny::tags$span(
    "Rolling vol window (days)",
    mod_tooltip(
      trigger  = bsicons::bs_icon("info-circle"),
      type     = "bslib",
      contents = "Number of trading days used for the rolling volatility calculation.",
      size     = "0.85rem",
      style    = "color:#6c757d"
    )
  ),
  min = 5L, max = 120L, value = 30L, step = 5L
)
```

The tooltip icon sits *inside* the label tag — it travels with the label
wherever the input is rendered with no additional server logic.

---

### Back-end: `"bslib"`

`bslib::popover()` opens on click and renders `contents` as HTML inside a
Bootstrap 5 popover. Pass `title` to show a popover header; pass `placement`
(`"top"`, `"right"`, `"bottom"`, `"left"`) to control direction.

```{r}
# Minimal popover — just a body
mod_tooltip(
  type     = "bslib",
  contents = "Annualised log return = mean daily log return \u00d7 252."
)

# Popover with a title and HTML list body, attached to a value-box title
mod_tooltip(
  trigger  = bsicons::bs_icon("info-circle-fill", class = "ms-1 text-info"),
  type     = "bslib",
  contents = "<ul>
    <li>Ann. Return: 14.2%</li>
    <li>Ann. Vol: 22.1%</li>
    <li>Sharpe: 0.64</li>
  </ul>",
  title    = "AAPL \u2013 Performance Summary",
  placement = "left"
)
```

**Where it appears in the app:** the *bslib* demo tab — each value box has a
clickable info icon in its title.

---

### Back-end: `"shinyhelper"`

`shinyhelper::helper()` wraps any tag and appends a circled-? icon. Clicking
the icon opens a modal dialog. Set `type = "inline"` (the default used here)
to render `contents` as an HTML string directly inside the modal body. Other
`type` values include `"markdown"` (reads a `.md` file) and `"text"` (plain
text).

`shinyhelper::observe_helpers()` must appear **once** in the UI (it is placed
in the nav panel header in `mod_outputs_ui()`).

```{r}
# Basic inline modal — content is an HTML string
mod_tooltip(
  trigger  = shiny::tags$span(class = "fs-4 fw-bold", "AAPL"),
  type     = "shinyhelper",
  contents = "<b>AAPL</b><br>Ann. Return: 14.2%<br>Sharpe: 0.64",
  type     = "inline",     # forwarded to shinyhelper::helper()
  title    = "AAPL \u2013 Performance Summary",
  size     = "m"           # modal size: "s", "m", "l"
)
```

**Where it appears in the app:** the *shinyhelper* demo tab — each metric card
shows a circled-? that opens a performance summary modal.

---

### Back-end: `"prompter"`

`prompter::add_prompt()` attaches a CSS-only tooltip via `data-*` HTML
attributes. No JavaScript initialisation is required beyond calling
`prompter::use_prompt()` **once** in the UI (placed in the nav panel header in
`mod_outputs_ui()`). Because the tooltip is CSS-driven, it works even in static
HTML contexts.

Key extra args (forwarded via `...`):

| Arg | Values | Effect |
|---|---|---|
| `position` | `"top"`, `"right"`, `"bottom"`, `"left"` | Tooltip placement |
| `rounded` | `TRUE` / `FALSE` | Rounded tooltip corners |
| `bounce` | `TRUE` / `FALSE` | Bounce animation on hover |
| `arrow` | `TRUE` / `FALSE` | Show/hide the tip arrow |

```{r}
# Tooltip on a metric label — appears to the right on hover
mod_tooltip(
  trigger  = shiny::tags$span("Ann. Return"),
  type     = "prompter",
  contents = "Annualised log return for AAPL: 14.2%",
  position = "right"
)

# Rounded, no arrow, bounces on hover
mod_tooltip(
  trigger  = shiny::tags$span("Sharpe"),
  type     = "prompter",
  contents = "Sharpe ratio (zero risk-free rate): 0.64",
  position = "top",
  rounded  = TRUE,
  bounce   = TRUE,
  arrow    = FALSE
)
```

**Where it appears in the app:** the *prompter* demo tab — each metric row
(`Ann. Return`, `Ann. Vol`, `Sharpe`) has a prompter tooltip on its label.

---

### Back-end: `"shinyalert"`

`shinyalert` is a modal-alert library. Unlike the other back-ends, it does not
use a hover event — it fires when the user *clicks* the trigger. `mod_tooltip()`
handles this by:

1. Calling `shinyalert::useShinyalert()` to inject the required JS dependency.
2. Attaching an `onclick` attribute to the trigger that calls the `shinyalert()`
   JavaScript function directly — no `shiny::observeEvent()` or server-side
   handler is needed.

The result is a `shiny::tagList()` of the dependency tag and the wired trigger.

Key extra args (forwarded and extracted via `...`):

| Arg | Description |
|---|---|
| `title` | Modal title bar text |
| `type` | Alert style: `"info"`, `"success"`, `"warning"`, `"error"`, `""` |
| `confirmButtonText` | Label for the confirm/close button |
| `closeOnClickOutside` | `TRUE` to dismiss by clicking outside the modal |

```{r}
# Clickable span that opens an info modal
mod_tooltip(
  trigger           = shiny::tags$span(class = "fs-4 fw-bold", "AAPL"),
  type              = "shinyalert",
  contents          = "Ann. Return: 14.2%\nAnn. Vol: 22.1%\nSharpe: 0.64",
  title             = "AAPL \u2013 Performance Summary",
  type              = "info",
  confirmButtonText = "Close",
  closeOnClickOutside = TRUE
)
```

**Where it appears in the app:** the *shinyalert* demo tab — clicking any
ticker card opens a formatted modal with full performance metrics.

---

## `mod_hoverinfo()`

### Signature

```{r}
mod_hoverinfo(
  type     = c("reactable", "gt"),
  contents = character(0),
  size     = NULL,
  style    = NULL,
  ...
)
```

### Parameters

| Parameter | Type | Description |
|---|---|---|
| `type` | character | Back-end: `"reactable"` (default) or `"gt"`. |
| `contents` | character vector | Hover content. Named vectors produce `"Name: value"` lines. See per-back-end notes below. |
| `size` | character | CSS font-size for the `"reactable"` span wrapper. Ignored by `"gt"`. |
| `style` | character | Inline CSS for the `"reactable"` span wrapper. Ignored by `"gt"`. |
| `...` | — | For `"reactable"`: extra HTML attributes on the `<span>`; for `"gt"`: passed to `gt::tab_footnote()`. |

### How it works — by back-end

#### `"reactable"`

Returns a single character string. Named vectors produce `"Name: value"` lines
joined by `sep` (`"\n"` by default). The string is assigned to the `text`
aesthetic in `ggplot2` and read by `plotly::ggplotly(tooltip = "text")`.

`mapply()` with `SIMPLIFY = TRUE` vectorises the call row-by-row across the
entire data frame, producing one string per data point:

```{r}
ggplot2::aes(
  text = mapply(
    function(sym, dt, adj) {
      mod_hoverinfo(
        type     = "plotly",
        contents = c(
          Ticker      = sym,
          Date        = as.character(dt),
          "Adj Close" = paste0("$", round(adj, 2))
        )
      )
    },
    sym = symbol, dt = date, adj = adjusted,
    SIMPLIFY = TRUE
  )
)
# Produces per-point strings like: "Ticker: AAPL\nDate: 2024-01-15\nAdj Close: $182.30"
```

#### `"reactable"`

Returns an `htmltools` `<span>` tag. `contents` is collapsed into a `title`
attribute (browser-native tooltip on hover), with named elements joined as
`"Name: value"` pairs separated by `" | "`. `size` and `style` become inline
CSS on the span. Additional child nodes (display text) pass through `...`:

```{r}
reactable::colDef(
  name = "Ann. Return (%)",
  cell = function(value, index) {
    mod_hoverinfo(
      type     = "reactable",
      contents = glue::glue(
        "Annualised log return for {df$symbol[index]}: {value}%"
      ),
      style    = paste0(
        "color:", if (value >= 0) "#198754" else "#dc3545",
        "; cursor:help"
      ),
      glue::glue("{value}%")  # display text passed via ...
    )
  }
)
```

The display text (`"{value}%"`) is an extra child node of the `<span>`, so the
cell renders the formatted value while the full hover description lives in
`title`.

#### `"gt"`

Returns a **list of argument lists** — one per named element of `contents`.
Each list contains `footnote` (text) and `locations` (a
`gt::cells_column_labels()` call). Apply them with `do.call()` in a loop:

```{r}
footnotes <- mod_hoverinfo(
  type     = "gt",
  contents = c(
    ann_return = "Annualised log return = mean daily log return \u00d7 252.",
    ann_vol    = "Annualised volatility = SD of daily log returns \u00d7 \u221a252.",
    sharpe     = "Sharpe ratio assumes a zero risk-free rate."
  )
)

tbl <- gt::gt(df)
for (fn_args in footnotes) {
  tbl <- do.call(gt::tab_footnote, c(list(tbl), fn_args))
}
```

`contents` **must** be a named character vector. Names are column identifiers
passed to `gt::cells_column_labels()` and must match actual column names in the
gt table.

---

## Decision guide

| Where does the tooltip appear? | Package / context | Function | `type` |
|---|---|---|---|
| Input label, heading, or icon in sidebar/UI | bslib | `mod_tooltip()` | `"bslib"` |
| Metric card with inline help modal | shinyhelper | `mod_tooltip()` | `"shinyhelper"` |
| Metric label with CSS-driven attribute tooltip | prompter | `mod_tooltip()` | `"prompter"` |
| Clickable element that opens a modal alert | shinyalert | `mod_tooltip()` | `"shinyalert"` |
| Cell tooltip in a reactable table | reactable | `mod_hoverinfo()` | `"reactable"` |
| Column-header footnote in a gt table | gt | `mod_hoverinfo()` | `"gt"` |

---

## Placement in the module stack

```
app_ui()
  └─ mod_inputs_ui()
       └─ mod_tooltip(type = "bslib")        <- label popovers in sidebar

  └─ mod_outputs_ui()
       └─ use_prompt()                        <- prompter CSS dependency (once)

app_server()
  └─ mod_inputs_server()                      <- returns reactive inputs list

  └─ mod_outputs_server()
       ├─ observe_helpers()                   <- shinyhelper server observer (once)
       ├─ renderUI  (bslib tab)
       │    └─ mod_tooltip(type = "bslib")
       ├─ renderUI  (shinyhelper tab)
       │    └─ mod_tooltip(type = "shinyhelper")
       ├─ renderUI  (prompter tab)
       │    └─ mod_tooltip(type = "prompter")
       ├─ renderUI  (shinyalert tab)
       │    └─ mod_tooltip(type = "shinyalert")
       ├─ renderReactable
       │    └─ mod_hoverinfo(type = "reactable")
       └─ render_gt
            └─ mod_hoverinfo(type = "gt")

  └─ mod_download_server()                    <- consumes perf_r
```

Neither helper requires its own `moduleServer()` call or namespace — they are
plain functions that slot into the existing module hierarchy at exactly the
point where tooltip content is needed.
