---
title: "Tooltip and Hover-Info Modules"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tooltip and Hover-Info Modules}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = FALSE
)
```

Two helper functions — `mod_tooltip()` and `mod_hoverinfo()` — give every
module in **tooltipexplorer** a consistent, package-agnostic way to attach
tooltip and hover-info content to UI elements and outputs.

- **`mod_tooltip()`** is a *UI helper*. It returns a `shiny.tag` and can be
  placed anywhere in a UI tree or inside another module's UI function — no
  server-side counterpart is needed.
- **`mod_hoverinfo()`** is a *server/logic helper*. It formats hover content
  for back-ends that build their tooltips programmatically (plotly, reactable,
  gt) rather than through HTML attributes.

The two functions cover complementary ground:

| Question | Function |
|---|---|
| "I want a hover tooltip on a *label* or *icon* in the UI" | `mod_tooltip()` |
| "I want hover content inside a *chart or table cell*" | `mod_hoverinfo()` |

---

## `mod_tooltip()`

### Signature

```{r}
mod_tooltip(
  trigger  = bsicons::bs_icon("info-circle"),
  type     = c("tippy", "bslib", "bootstrap"),
  contents = "",
  size     = NULL,
  style    = NULL,
  ...
)
```

### Parameters

| Parameter | Type | Description |
|---|---|---|
| `trigger` | `shiny.tag` or text | The element the tooltip attaches to. Defaults to a small info-circle icon. |
| `type` | character | Back-end: `"tippy"` (default), `"bslib"`, or `"bootstrap"`. |
| `contents` | character | Tooltip body text. HTML is accepted for `"tippy"` and `"bslib"`. |
| `size` | character | CSS font-size for the trigger wrapper, e.g. `"0.85rem"`. `NULL` leaves it unchanged. |
| `style` | character | Additional inline CSS for the trigger wrapper, e.g. `"color:#6c757d"`. `NULL` applies none. |
| `...` | — | Extra arguments forwarded to the underlying back-end (`tippy::tippy()`, `bslib::popover()`). |

### How it works

`size` and `style` are combined into a single inline CSS string and applied to
a `<span>` that wraps `trigger`. The wrapped trigger is then passed to the
chosen back-end:

- **`"tippy"`** — calls `tippy::tippy()` with `allowHTML = TRUE`. Any `...`
  args (e.g. `placement`, `theme`, `animation`) go straight to
  `tippy::tippy()`.
- **`"bslib"`** — calls `bslib::popover()`, wrapping `contents` in
  `shiny::HTML()` so HTML markup renders. Any `...` args (e.g. `title`,
  `placement`) go to `bslib::popover()`.
- **`"bootstrap"`** — attaches `data-bs-toggle="tooltip"`,
  `data-bs-title`, and `title` attributes directly to the trigger tag. No
  external JS is injected; Bootstrap's tooltip JavaScript (already loaded by
  bslib) picks these up automatically.

### Usage in a module UI

```{r}
# Inside mod_inputs_ui() — tooltip on a slider label
shiny::sliderInput(
  inputId = ns("vol_window"),
  label   = shiny::tags$span(
    "Rolling vol window (days)",
    mod_tooltip(
      trigger  = bsicons::bs_icon("info-circle"),
      type     = "tippy",
      contents = "Number of trading days used for the rolling volatility calculation.",
      size     = "0.85rem",
      style    = "color:#6c757d"
    )
  ),
  min = 5L, max = 120L, value = 30L, step = 5L
)
```

The tooltip icon sits *inside* the label tag, so it travels with the label
wherever the input is rendered — no additional server logic required.

### Back-end examples

#### tippy

```{r}
# Plain text
mod_tooltip(
  trigger  = shiny::tags$span("Hover me"),
  type     = "tippy",
  contents = "Hello from tippy!"
)

# HTML contents + extra tippy options
mod_tooltip(
  trigger   = bsicons::bs_icon("question-circle"),
  type      = "tippy",
  contents  = "<b>Sharpe ratio</b><br>Assumes a zero risk-free rate.",
  size      = "0.8rem",
  style     = "color:#0d6efd; cursor:help",
  placement = "right",
  theme     = "light-border"
)
```

#### bslib

```{r}
# Popover with an HTML list body and a title
mod_tooltip(
  trigger  = bsicons::bs_icon("info-circle-fill", class = "ms-1 text-info"),
  type     = "bslib",
  contents = "<ul>
    <li>Ann. Return: 14.2%</li>
    <li>Ann. Vol: 22.1%</li>
    <li>Sharpe: 0.64</li>
  </ul>",
  title    = "AAPL — Performance Summary"
)
```

#### bootstrap

```{r}
# Adds data-bs-toggle="tooltip" attributes; Bootstrap JS activates on hover
mod_tooltip(
  trigger  = shiny::tags$span("ⓘ", class = "text-muted"),
  type     = "bootstrap",
  contents = "Annualised log return × 252 trading days."
)
```

---

## `mod_hoverinfo()`

### Signature

```{r}
mod_hoverinfo(
  type     = c("plotly", "reactable", "gt"),
  contents = character(0),
  size     = NULL,
  style    = NULL,
  sep      = "\n",
  ...
)
```

### Parameters

| Parameter | Type | Description |
|---|---|---|
| `type` | character | Back-end: `"plotly"` (default), `"reactable"`, or `"gt"`. |
| `contents` | character vector | Hover content. Named vectors produce `"Name: value"` lines. See per-back-end notes below. |
| `size` | character | CSS font-size for the `"reactable"` span wrapper. Ignored by `"plotly"` and `"gt"`. |
| `style` | character | Inline CSS for the `"reactable"` span wrapper. Ignored by `"plotly"` and `"gt"`. |
| `sep` | character | Line separator for `"plotly"` output. Defaults to `"\n"`. |
| `...` | — | Forwarded to the underlying helper. For `"reactable"`, passed as extra HTML attributes on the `<span>`; for `"gt"`, passed to `gt::tab_footnote()`. |

### How it works — by back-end

#### `"plotly"`

Returns a single character string. Named vectors produce `"Name: value"` lines
joined by `sep` (`"\n"` by default). This string is assigned to the `text`
aesthetic in `ggplot2` and then read by `plotly::ggplotly(tooltip = "text")`.

```{r}
# Inside mod_outputs_server() — building the plotly text aesthetic
ggplot2::aes(
  text = mapply(
    function(sym, dt, adj) {
      mod_hoverinfo(
        type     = "plotly",
        contents = c(
          Ticker      = sym,
          Date        = as.character(dt),
          "Adj Close" = paste0("$", round(adj, 2))
        )
      )
    },
    sym = symbol, dt = date, adj = adjusted,
    SIMPLIFY = TRUE
  )
)
# Each point's hover text: "Ticker: AAPL\nDate: 2024-01-15\nAdj Close: $182.30"
```

`mapply()` with `SIMPLIFY = TRUE` vectorises the call across all rows of the
data frame, producing one string per observation.

#### `"reactable"`

Returns an `htmltools` `<span>` tag. The `contents` vector is collapsed into a
`title` attribute (using `" | "` as a separator for multi-element vectors),
which browsers display as a native tooltip on hover. `size` and `style` are
applied as inline CSS on the span.

Pass the return value directly as the `cell` renderer in
`reactable::colDef()`:

```{r}
reactable::colDef(
  name = "Ann. Return (%)",
  cell = function(value, index) {
    mod_hoverinfo(
      type     = "reactable",
      contents = glue::glue(
        "Annualised log return for {df$symbol[index]}: {value}%"
      ),
      style    = paste0(
        "color:", if (value >= 0) "#198754" else "#dc3545",
        "; cursor:help"
      ),
      glue::glue("{value}%")   # <-- display text passed via ...
    )
  }
)
```

The display text (`"{value}%"`) is passed through `...` as an additional child
node of the `<span>`, so the cell renders the formatted value while the full
hover string lives in `title`.

#### `"gt"`

Returns a **list of argument lists** — one per named element of `contents`.
Each list has two elements:

- `footnote` — the footnote text string
- `locations` — a `gt::cells_column_labels()` call targeting that column

These are applied to a gt table by iterating with `do.call(gt::tab_footnote, .)`:

```{r}
# Build footnote args for three columns at once
footnotes <- mod_hoverinfo(
  type     = "gt",
  contents = c(
    ann_return = "Annualised log return = mean daily log return × 252.",
    ann_vol    = "Annualised volatility = SD of daily log returns × √252.",
    sharpe     = "Sharpe ratio assumes a zero risk-free rate."
  )
)

# Apply each footnote in a loop
tbl <- gt::gt(df)
for (fn_args in footnotes) {
  tbl <- do.call(gt::tab_footnote, c(list(tbl), fn_args))
}
```

`contents` **must** be a named character vector for `"gt"`. The names are
column identifiers passed to `gt::cells_column_labels(columns = col)` and must
match actual column names in the gt table.

---

## Decision guide

Use this table to pick the right function and `type` for any given situation:

| Where does the tooltip appear? | Package / context | Function | `type` |
|---|---|---|---|
| Input label, heading, icon in sidebar or UI | Any | `mod_tooltip()` | `"tippy"` |
| Value box title or showcase icon | bslib | `mod_tooltip()` | `"bslib"` |
| Static span or button in UI | Bootstrap | `mod_tooltip()` | `"bootstrap"` |
| Hover text on a ggplot → plotly chart | plotly | `mod_hoverinfo()` | `"plotly"` |
| Cell tooltip in a reactable table | reactable | `mod_hoverinfo()` | `"reactable"` |
| Column-header footnote in a gt table | gt | `mod_hoverinfo()` | `"gt"` |

## Placement in the module stack

```
app_ui()
  └─ mod_inputs_ui()          ← mod_tooltip() used in labels
  └─ mod_outputs_ui()         ← output placeholders only

app_server()
  └─ mod_inputs_server()      ← returns reactive inputs list
  └─ mod_outputs_server()     ← mod_hoverinfo() used in render* functions
       └─ renderPlotly         mod_hoverinfo(type = "plotly",  ...)
       └─ renderReactable      mod_hoverinfo(type = "reactable", ...)
       └─ render_gt            mod_hoverinfo(type = "gt", ...)
  └─ mod_download_server()    ← consumes perf_r from mod_outputs_server()
```

Neither helper requires its own `moduleServer()` call or namespace — they are
plain functions that fit into the existing module hierarchy at exactly the
point where tooltip content is needed.
